Project Description -- Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media

### 2019-11-12 docs/proposals/NSF/ProjDesc/ProjDesc.tex
NM

% Removed this paragraph because it emphasized sinkholes too much.
%To take one example (of particular relevance in the state of Florida) the formation of sinkholes can be triggered by certain human activities such as groundwater pumping or the weight of run-off storage ponds \cite{sandhu2018fate}. When collapse occurs below a storage pond, the underlying reservoir of pristine groundwater can be severely contaminated, thus surrendering a natural resource of great economic and societal value. Clearly, a deeper understanding of the mechanisms leading to sinkhole formation, and more generally that underly earth's natural filtration system, could reap enormous benefits in predicting and mitigating such environmental disasters. Other applications include understanding and predicting contamination by nuclear waste and by hydraulic fracturing.

@@

%---------------------------------------------------------------------------------------------------%
\subsection{Transport through eroded media}
\begin{wrapfigure}[]{r}{0.3\textwidth}
  \includegraphics[width=0.3\textwidth]{figs/100b_t100tracer}
  \caption{\label{fig:100tracers} A visualization of the streamlines
  with 200 tracers passing through an eroded geometry with a porosity of
  0.63. The tracers are initialized at the left end of the channel.}
\end{wrapfigure}
Transport through eroded media is a direct consequence of the pore
structure. Qualitative properties of an eroded media, such as channels
of high porosity can be modest in volume fraction, but still be
responsible for a large portion of the flux~\cite{Quaife2018}. To
increase understanding of flow through eroded media, we will simulate
and statistically analyze transport through eroded geometries.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%  SECTION: Education and outreach
%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Broader impacts of the proposed work}

By gaining a deeper understanding of the underlying physical processes, this investigation offers several potential benefits for managing water resources.
\nick{More on erosion modifying porous medium properties etc...}
%First and foremost, a better understanding of sinkhole collapse could lead to policies to prevent these natural hazards. Second, inference of spatially-variable porous-medium properties from seepage measurements could identify locations in natural aquifers vulnerable to contamination and/or collapse.

The above scientific impacts have been addressed throughout this proposal, and so the section below expands on how this project will impact future generations of mathematical scientists through educational and outreach activities, particularly women and underrepresented minorities (URM). 
The PIs are already actively involved in graduate and undergraduate research, as well as outreach to K-12 students. Further, we have successfully recruited women and underrepresented minorities in both undergraduate and graduate research. The section below describes a range of new educational activities related to the proposed research.

\subsection{Undergraduate research}
The ideas in this proposal represent a rich and fascinating topic for undergraduate research, while also being approachable by the methods and tools acquired at that stage. As mentioned above, there are a number of open questions in this proposal appropriate for undergraduate research, from {\color{blue} Stuff}

At FSU, there are at least two excellent avenues to attract interested undergraduate students:  the Undergraduate Research Opportunity Program (UROP) and the IDEA grant. The UROP is a program to engage underclassmen in academic research, while the IDEA grant is a competitive program, requiring (typically advanced) students to write a proposal in which they identify a research advisor. Both programs make special efforts to recruit from underrepresented groups. I already have experience using these programs to recruit interested students as I am currently serving as the advisor for IDEA grant winner Tyler Bolles, whose proposal was ranked $\#1$ overall out of the 110 applicants.

I will use a combination of funds from the above two resources and from the current proposal to engage undergraduates in research. Since the UROP program is well-known at FSU, it is the best way to recruit young, interested students. I plan to recruit one such student who will be supported by UROP for the first summer of his/her research, and should the student wish to continue, I will fund subsequent years using the current proposal. I will also encourage the student to apply for the IDEA grant in the final year of research, as this is a prestigious award at FSU.

\subsection{Assessment for graduate and undergraduate education}

Developing strong mentoring relationships is essential to a student's intellectual development and a responsibility that the PIs takes very seriously. When working with students on research, whether undergraduate or graduate, we are always sure to outline the expectations of {\em both} parties -- the student and myself -- and periodically revisit the list to determine if the expectations are being met or if changes are necessary. This ensures that progress continues uninterrupted and that both parties are comfortable with their responsibilities. For the graduate seminar course, questionnaires will be provided to the students to determine how the seminar relates to their own research and their intellectual development, as well as parts of the course that were particularly beneficial and parts that could be improved. This will ensure that, once integrated into the standard curriculum, the course will maximally benefit the students' education.

@@

Plans for future outreach include continued collaboration with CPALMS to create a series of educational videos for both students and teachers, as well as planned activities at {\bf Math Fun Day}, which is an annual event held by the FSU Mathematics departments to engage K-12 students in the region. Moore has led activities at Math Fun Day in prior years and, related to the current proposal, plans to deliver lectures and demos on porous-media flows, erosion, and sinkhole collapse.
Laboratory materials for the demo are available at the neighboring Geophysical Fluid Dynamics Institute, of which the PI is a faculty associate.

>>>>

% REMOVED this paragraph because it emphasized sinkholes too much.
%To take one example (of particular relevance in the state of Florida) the formation of sinkholes can be triggered by certain human activities such as groundwater pumping or the weight of run-off storage ponds \cite{sandhu2018fate}. When collapse occurs below a storage pond, the underlying reservoir of pristine groundwater can be severely contaminated, thus surrendering a natural resource of great economic and societal value. Clearly, a deeper understanding of the mechanisms leading to sinkhole formation, and more generally that underly earth's natural filtration system, could reap enormous benefits in predicting and mitigating such environmental disasters. Other applications include understanding and predicting contamination by nuclear waste and by hydraulic fracturing.

@@

%---------------------------------------------------------------------------------------------------%
\subsection{Transport through eroded media}

\begin{wrapfigure}[]{r}{0.3\textwidth}
  \includegraphics[width=0.3\textwidth]{figs/100b_t100tracer}
  \caption{\label{fig:100tracers} A visualization of the streamlines
  with 200 tracers passing through an eroded geometry with a porosity of
  0.63. The tracers are initialized at the left end of the channel.}
\end{wrapfigure}

Transport through eroded media is a direct consequence of the pore
structure. Qualitative properties of an eroded media, such as channels
of high porosity can be modest in volume fraction, but still be
responsible for a large portion of the flux~\cite{Quaife2018}. To
increase understanding of flow through eroded media, we will simulate
and statistically analyze transport through eroded geometries.

@@

%---------------------------------------------------------------------------------------------------%
%%%%  SECTION: Education and outreach
%---------------------------------------------------------------------------------------------------%
\section{Broader impacts of the proposed work}

	This research proposed here could lead to a range of societal benefits, including better management of water resources, accurate prediction of contaminant transport, and an understanding of the long-term effects of human activities, such as hydraulic fracturing or groundwater pumping, on porous media and water resources. In such scenarios, the slow timescales of mechanical and chemical erosion are coupled to the faster timescales of human-induced changes and/or gravitational collapse. Currently, the relationship between these scales is poorly understood. The high-fidelity, efficient computational methods proposed here will enable a deeper understanding of these processes, which is the first step towards developing effective strategies for managing water resources and mitigating catastrophic events such as contamination or structural collapse. For example, the insights gained here could potentially help identify locations in natural aquifers vulnerable to contamination and/or collapse.

	In addition to these societal impacts, the work proposed here will spur the training of mathematicians and computational scientists in fundamentally cross-disciplinary research of great societal importance. Further, the PIs will engage in educational and outreach activities to impact future generations of mathematical scientists, with particular focus on women and underrepresented minorities (URM). 
Both PIs are already actively involved in graduate and undergraduate research, as well as outreach to K-12 students. Further, we have successfully recruited women and underrepresented minorities in undergraduate, graduate, and postdoctoral research.

\subsection{Graduate research}

	An important benefit of the research proposed here is the potential to immerse graduate students in fundamentally {\em cross-disciplinary} research. This proposal involves a confluence of disciplines, most centrally computational mathematics, but also geophysics, reduced-physical modeling, hydrology, and environmental science. Exposure to a range of disciplines is, in the view of the PIs, an invaluable opportunity for young scientists.

	Naturally, the research proposed here will form the foundation for the PhD dissertations of the graduate students involved. Beyond this, the students will engage in publishing articles in peer-reviewed journals of high quality and attending national and international conferences at which they will contribute presentations on their work. The PIs will closely mentor the graduate students to ensure their continued development, not only intellectually, but also in the `soft' skills of communication, promotion, and networking. Having both PIs Moore and Quaife involved in the training of each graduate student is especially valuable for offering multiple perspectives on these issues. The PIs will make special efforts to recruit women and underrepresented minorities as the graduate students funded by this project. PI Moore has already had success recruiting women in these roles, as the first graduate student to receive her PhD under Moore's advising was female (Karina Khazmutdinova, 2016).

\subsection{Undergraduate research}

	Regarding undergraduate research, the ideas in this proposal represent a rich and fascinating topic for undergraduate students, while also being approachable these students. In particular, we aim to recruit at least one undergraduate student to perform laboratory experiments at the Geophysical Fluid Dynamics Institute (GFDI). As mentioned above, GFDI offers state-of-the-art laboratory facilities, with an already-functioning sinkhole experiment. The student will perform controlled experiments using natural and artificial porous media (e.g.~sand and micro-glass beads respectively) to precisely characterize the parameter regime leading to gravitational collapse. Parallel experiments will be conducted with erodible (bentonite clay) or dissolvable  (sucrose solids) fragments materials, which will represent the porous medium. These experiments will be used to guide, calibrate, and validate the computational advances. These undergraduate students will work closely with the graduate students to ensure that theory and experiments advance in close cooperation, with constant feedback between the two.

	PI Moore already has experience successfully training undergraduate students in combined laboratory and theoretical work. A recent example is undergraduate student Tyler Bolles who completed an honors thesis with PI Moore at FSU on laboratory and theory of water waves. This work resulted in one {\em peer-reviewed} publication already \cite{Bolles2019}, with a second to be submitted soon.

	At FSU, there are at least two excellent avenues to attract interested undergraduate students:  the Undergraduate Research Opportunity Program (UROP) and the IDEA grant. The UROP is a program to engage underclassmen in academic research, while the IDEA grant is a competitive program, requiring (typically advanced) students to write a proposal in which they identify a research advisor. Both programs make special efforts to recruit from underrepresented groups. PIs Moore and Quaife already have experience using these programs to recruit interested students.

@@

Plans for future outreach include continued collaboration with CPALMS to create a series of educational videos for both students and teachers, as well as planned activities at {\bf Math Fun Day}, which is an annual event held by the FSU Mathematics departments to engage K-12 students in the region. PI Moore has led activities at Math Fun Day in prior years and, related to the current proposal, plans to deliver lectures and demos on porous-media flows, erosion, and sinkhole collapse. Laboratory materials for the demo are available at GFDI.

%\subsection{Assessment for graduate and undergraduate education}
%Developing strong mentoring relationships is essential to a student's intellectual development and a responsibility that the PIs takes very seriously. When working with students on research, whether undergraduate or graduate, we are always sure to outline the expectations of {\em both} parties -- the student and myself -- and periodically revisit the list to determine if the expectations are being met or if changes are necessary. This ensures that progress continues uninterrupted and that both parties are comfortable with their responsibilities. For the graduate seminar course, questionnaires will be provided to the students to determine how the seminar relates to their own research and their intellectual development, as well as parts of the course that were particularly beneficial and parts that could be improved. This will ensure that, once integrated into the standard curriculum, the course will maximally benefit the students' education.
